Smoothing Algorithms for Variable Rate Models

### 2012-04-06 draft2_VRsmoothing.tex
Still polishing

\author{Pete~Bunch,~\IEEEmembership{}
        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{P. Bunch and S. Godsill are with the Department
of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}

@@

\begin{abstract}
The abstract goes here.
\end{abstract}

@@

\IEEEPARstart{T}{he} objective of sequential Bayesian inference is to estimate an imperfectly observed quantity as it varies over time. This is accomplished through the use of probabilistic models for the state evolution and measurement processes. Often, the latent state is a continuously varying quantity, whereas the observations are made at a discrete set of times. In these circumstances, it is simplest to discretise the state onto the same set of times as the observations. When the system is also Markovian, this leads to the standard ``fixed rate'' hidden Markov model (HMM). The standard HMM is poorly suited to systems where the state evolution contains discontinuities; for example, the price of a financial asset which may display large jumps at random times between periods of diffusion-like behaviour, or the kinematic state of a manoeuvring vehicle which may have sudden changes in the acceleration when turns begin or end. Such problems can be handled more naturally using a ``variable rate'' model, in which the state dynamics are conditioned upon a set of random changepoints which characterise transitions in behaviour.

In a variable rate model, the set of changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are thoroughly set out in \cite{Jacobsen2006}. Conditional upon this MPP the state evolves according to some benign dynamics. In \cite{Godsill2007,Whiteley2011}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.

The posterior distribution for the changepoint MPP is inherently nonlinear, and cannot be calculated analytically. Instead, inference must be conducted using numerical approximations. The particle filter (introduced by \cite{Gordon1993}) and smoother (see \cite{Doucet2000a,Godsill2004}) are schemes which approximate a posterior distribution using a set of samples, or ``particles'', drawn sequentially from it.  A thorough introduction to particle filtering and smoothing methods can be found in \cite{Cappe2007,Doucet2009}. In \cite{Godsill2007a,Godsill2007,Whiteley2011}, the particle filter was adapted for use with variable rate models, resulting in the variable rate particle filter (VRPF).

@@

We introduce the general structure of variable rate models in section~\ref{sec:vr_models} and discuss two particular types, those with conditionally linear-Gaussian and conditionally deterministic dynamics. The VRPF is reviewed in section~\ref{sec:vrpf}, and new variable rate smoothing algorithms are described in section~\ref{sec:vrps}. In section~\ref{sec:simuations}, particular examples of variable rate models are presented and their performance demonstrated in a series of simulations.

@@

We consider a general model from time $0$ to $T$, between which observations, $\{y_1 \dots y_N\}$, are made at times $\{t_1 \dots t_N = T\}$. During this period, an unknown number of changepoints, $K$, occur at times $\{\tau_0 = 0, \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_0, u_1 \dots u_K \}$. The pairs $\{\tau_k, u_k\}$ are the elements of an marked point process (MPP). We will refer to the times of the MPP as the ``change-time sequence'' and marks as ``parameter sequence'', and both together as the ``changepoint sequence''. The latent state is a continuous-time process denoted $x(t)$. Discrete sets containing multiple values over time will be written as, e.g. $y_{1:n} = \{y_1 \dots y_n\}$.

@@

The changepoint sequence is assumed to be a Markov process.
%
\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(u_k|\tau_k, \tau_{k-1}, u_{k-1}) p(\tau_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
\end{IEEEeqnarray}

The changepoint density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.

@@

By introducing a new sequence, $\{ x_0, x_1 \dots x_K \}$, which denotes the value of the state at each changepoint (i.e. $x(\tau_k)$), and assuming that an analytic solution exists, a state transition function may be found,
%
\begin{IEEEeqnarray}{rCll}
 x(t) & = & f(x_{K_n}, v_{K_n}, \tau_{K_n}, t) &, \tau_{K_n} < t <= \tau_{K_{n}+1}    \label{eq:disc_time_state_diff_eq}     .
\end{IEEEeqnarray}

By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times may be calculated. These points will be denoted $\hat{x}_n$. To complete the model, a probabilistic measurement model must be chosen for the observation process, $p(y_n|\hat{x}_n)$.

For convenience, we assume that $x_0$ is known in the following sections. This means that $x(t)$ may be calculated deterministically for all $t$ given $\theta$. However, this condition is easily relaxed by including $x_0$ as a random variable in the posterior distribution.

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.

@@

The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a}. The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$. This distribution may be expanded using Bayes' rule.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|y_{1:n})} \nonumber \\
 & \propto & p(y_n|\theta_{n}, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}
\end{IEEEeqnarray}

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, can be constructed in a similar manner to (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, as a product of density terms for each new changepoint and a survivor function term. In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
\end{IEEEeqnarray}

For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}). For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that a changepoint cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
\end{IEEEeqnarray}

@@

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is proposed from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated to $\theta_{n-1}$ to create an estimate of $\theta_n$. Finally, the particle is weighted according to the ratio of the target and proposal densities.
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}
\end{IEEEeqnarray}

@@

The choice of proposal weights, $\{v_{n-1}^{(i)}\}$, requires particular attention in the design of VRPFs. In some models a changepoint may not have an immediate effect on the observations, especially if a jump occurs in some quantity which is only observed via its integral, e.g. if there is a jump in the acceleration of a moving object, yet only the position is measured, the change will not be apparent until several more observations have been made. In the meantime, particles which contain a changepoint at the correct time may all have been removed by the resampling process. To avoid this loss of good particles, proposal weights should be chosen which preserve a significant number of low-weight particles. One scheme which has been found to work well is described in \cite{Godsill2007}, in which proposal weights are given by:
%
\begin{IEEEeqnarray}{rCl}
v_{n-1}^{(i)} & \propto & \max ( 1, N_F w_{n-1}^{(i)} )
\end{IEEEeqnarray}

@@

For a conditionally linear-Gaussian model, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Anderson1979},
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

@@

This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}. The concept of estimating the posterior for just the changepoint sequence with a particle filter rather than the joint posterior over changepoints and linear states is an example of Rao-Blackwellisation (see, e.g. \cite{Casella1996,Doucet2000}), so this algorithm is known as the Rao-Blackwellised variable rate particle filter (RB-VRPF) \cite{Godsill2007a,Christensen2012}.

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}$, where $\tau_0^{(i)}=0$ and $\{u_0^{(i)}\} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a changepoint history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose a sequence extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
    \STATE Add extension to changepoint sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict state mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}. The resulting trajectories $x(t)^{(i)}$ for each particle are realisations from a piecewise-deterministic process, so this algorithm is named the piecewise-deterministic variable rate particle filter (PD-VRPF).

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}$, where $\tau_0^{(i)}=0$ and $\{u_0^{(i)}\} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose a sequence extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
    \STATE Add extension to changepoint sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate particle weight using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Piecewise deterministic variable rate particle filter}
\label{alg:VRPF}
\end{figure}

@@

The bootstrap versions of the VRPFs described in sections~\ref{sec:rb-vrpf} and~\ref{sec:pd-vrpf} may perform poorly if changepoints are not obvious until significantly after they occur. For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of observations have arrived. In this case, the estimation may be improved by the introduction of resample-move (RM) steps \cite{Gilks2001}. In an RM scheme, optional Metropolis-Hastings (MH) moves are conducted to alter the particle states after the importance sampling has taken place. For variable rate models, any one of the previous changepoints, $\tau_k$, or associated parameters, $u_k$, could be adjusted. Because more observations are available than when the changepoint was first proposed, it may be possible to construct more informed proposals and so move the changepoints towards regions with higher posterior probability. It is even possible to retrospectively add or remove changepoints, using reversible jump MH moves \cite{Green1995}. Variable rate particle filters using RM with piecewise deterministic models are described in \cite{Whiteley2011,Gilholm2008}.

@@

Filtering schemes which alter past changepoints -- whether using an SMC sampler or RM -- are computationally expensive, because many likelihood calculations must be conducted, for each observation from the time of the change onwards. In some cases, it may be simpler to just use a bootstrap filter with more particles.

@@

A filter conducts inference sequentially as new observations are introduced. The purpose of a smoother is to produce a second estimate once all the observations have been made, using future values to improve upon the filter performance. Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened. Thus, it is anticipated that a smoothing algorithm will provide significantly improved performance at changepoint estimation.

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. A particle approximation to this distribution is generated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity -- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a good characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.

@@

The transition density term will include a delta function with unit probability mass at the point for which the past and future state trajectories match up.

>>>>

\author{Pete~Bunch,~\IEEEmembership{Member,~IEEE,}
        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{P. Bunch and S. Godsill are with the Department
of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}

@@

\begin{abstract}
The abstract goes here. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah. Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
\end{abstract}

@@

\IEEEPARstart{T}{he} objective of sequential Bayesian inference is to estimate an imperfectly observed quantity as it varies over time. This is accomplished through the use of probabilistic models for the state evolution and measurement processes. Often, the latent state is a continuously varying quantity, whereas the observations are made at a discrete set of times. In these circumstances, it is simplest to discretise the state onto the same set of times as the observations. When the system is also Markovian, this leads to the standard ``fixed rate'' hidden Markov model (HMM). The standard HMM is poorly suited to systems where the state evolution contains discontinuities; for example, the price of a financial asset which may display large jumps at random times between periods of diffusion-like behaviour, or the kinematic state of a manoeuvring vehicle which may have sudden changes in the acceleration when turns begin or end. Such problems can be handled more naturally using a ``variable rate'' model, in which the state dynamics are conditioned upon a set of changepoints which characterise transitions in behaviour.

In a variable rate model, the set of changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are set out thoroughly in \cite{Jacobsen2006}. Conditional upon this MPP the state evolves according to some benign dynamics. In \cite{Godsill2007,Whiteley2011}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.

The posterior distribution for the changepoint MPP is inherently nonlinear, and cannot be calculated analytically. Instead, inference may be conducted using numerical approximations. The particle filter (introduced by \cite{Gordon1993}) and smoother (see \cite{Doucet2000a,Godsill2004}) are schemes which approximate a posterior distribution using a set of samples, or ``particles'', drawn sequentially from it.  A thorough introduction to particle filtering and smoothing methods can be found in \cite{Cappe2007,Doucet2009}. In \cite{Godsill2007a,Godsill2007,Whiteley2011}, the particle filter was adapted for use with variable rate models, resulting in the variable rate particle filter (VRPF).

@@

We introduce the general structure of variable rate models in section~\ref{sec:vr_models}. The VRPF is reviewed in section~\ref{sec:vrpf}, and new variable rate smoothing algorithms are described in section~\ref{sec:vrps}. In section~\ref{sec:simulations}, particular examples of variable rate models are presented and algorithm performance is demonstrated in a series of simulations.

@@

We consider a general model from time $0$ to $T$, between which observations, $\{y_1 \dots y_N\}$, are made at times $\{t_1 \dots t_N = T\}$. During this period, an unknown number of changepoints, $K$, occur at times $\{\tau_0 = 0, \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_0, u_1 \dots u_K \}$. The pairs $\{\tau_k, u_k\}$ are the elements of an marked point process (MPP). The latent state is a continuous-time process denoted $x(t)$. Discrete sets containing multiple values over time will be written as, e.g. $y_{1:n} = \{y_1 \dots y_n\}$.

@@

The changepoint sequence is assumed to be a Markov process.
%
\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
\end{IEEEeqnarray}

This density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.

@@

By introducing a new sequence, $\{ x_0, x_1 \dots x_K \}$, which denotes the value of the state at each changepoint (i.e. $x(\tau_k)$), and assuming that an analytic solution exists, a state transition function may be found,
%
\begin{IEEEeqnarray}{rCll}
 x(t) & = & f(x_{K_n}, u_{K_n}, \tau_{K_n}, t) &, \tau_{K_n} < t \leq \tau_{K_{n}+1}    \label{eq:disc_time_state_diff_eq}     .
\end{IEEEeqnarray}

By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times may be calculated --- these points will be denoted $\hat{x}_n$. To complete the model, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.

For convenience, we assume that $x_0$ is known in the following sections. This means that $x(t)$ may be calculated deterministically for all $t$ given $\theta$. This condition is easily relaxed by including $x_0$ as a random variable in the posterior distribution.

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007,Whiteley2011}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.

@@

The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a,Whiteley2011}. The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$. This distribution may be expanded using Bayes' rule,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{ p(\theta_{n}|y_{1:n}) \propto p(y_n|\theta_{n}, y_{1:n-1}) } \nonumber \\
 \qquad & & \times p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}     .
\end{IEEEeqnarray}

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, comprising a product of density terms (one for each new changepoint) and a survivor function term.% In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
\end{IEEEeqnarray}

For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}). For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that it cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
\end{IEEEeqnarray}

@@

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is proposed from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}$ to create an estimate of $\theta_n$. Finally, the particle is weighted according to the ratio of the target and proposal densities.
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}
\end{IEEEeqnarray}

@@

%The choice of proposal weights, $\{v_{n-1}^{(i)}\}$, requires particular attention in the design of VRPFs. In some models a changepoint may not have an immediate effect on the observations, especially if a jump occurs in some quantity which is only observed via its integral, e.g. if there is a jump in the acceleration of a moving object, yet only the position is measured, the change will not be apparent until several more observations have been made. In the meantime, particles which contain a changepoint at the correct time may all have been removed by the resampling process. To avoid this loss of good particles, proposal weights should be chosen which preserve a significant number of low-weight particles. One scheme which has been found to work well is described in \cite{Godsill2007}, in which proposal weights are given by:

In \cite{Godsill2007}, the following choice of proposal weights was found to work well, as it preserves low probability particles which might turn out to be good estimates at later time steps,
%
\begin{IEEEeqnarray}{rCl}
v_{n-1}^{(i)} & \propto & \max ( 1, N_F w_{n-1}^{(i)} )     .
\end{IEEEeqnarray}

@@

For a conditionally linear-Gaussian model, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

@@

This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}.% The concept of estimating the posterior for just the changepoint sequence with a particle filter rather than the joint posterior over changepoints and linear states is an example of Rao-Blackwellisation (see, e.g. \cite{Casella1996,Doucet2000}), so this algorithm is known as the Rao-Blackwellised variable rate particle filter (RB-VRPF) \cite{Godsill2007a,Christensen2012}.

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.% The resulting trajectories $x(t)^{(i)}$ for each particle are realisations from a piecewise-deterministic process, so this algorithm is named the piecewise-deterministic variable rate particle filter (PD-VRPF).

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.7} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Piecewise deterministic variable rate particle filter}
\label{alg:VRPF}
\end{figure}

@@

The bootstrap versions of the VRPF may perform poorly if changepoints are not obvious until significantly after they occur. For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of observations have arrived. In this case, the estimation may be improved by the introduction of resample-move (RM) steps \cite{Gilks2001}. In an RM scheme, optional Metropolis-Hastings (MH) moves are conducted to alter the particle states after the importance sampling has taken place. For variable rate models, any one of the previous changepoints, $\tau_k$, or associated parameters, $u_k$, could be adjusted. Because more observations are available than when the changepoint was first proposed, it may be possible to construct more informed proposals and so move the changepoints towards regions with higher posterior probability. It is even possible to retrospectively add or remove changepoints, using reversible jump MH moves \cite{Green1995}. Variable rate particle filters using RM with piecewise deterministic models are described in \cite{Whiteley2011,Gilholm2008}.

@@

%Filtering schemes which alter past changepoints -- whether using an SMC sampler or RM -- are computationally expensive, because many likelihood calculations must be conducted, for each observation from the time of the change onwards. In some cases, it may be simpler to just use a bootstrap filter with more particles.

@@

A filter conducts inference sequentially as new observations are introduced. The purpose of a smoother is to produce a revised estimate once all the observations have been made, using future values to improve upon the filter performance. Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened. Thus, a smoothing algorithm is expected to achieve significantly improved performance at changepoint estimation.

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This is the distribution approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.

@@

%The transition density term will include a delta function with unit probability mass at the point for which the past and future state trajectories match up.
